A primer to numerical simulations: The perihelion motion of Mercury

### 2018-01-15 tex/MercuryNumerical.tex
added scripts url + tiny mods

Numerical simulations play a key role in modern physics for they allow one to tackle theoretical problems not accessible otherwise. This might be the case because there are too many particles participating in the system (as in simulations for weather predictions) or the interactions are too complicated to allow for a systematic, perturbative approach (as in theoretical descriptions of nuclear particles at the fundamental level).
This paper introduces a project that allows one to demonstrate the power of numerical simulations.
On the example of the perihelion motion of the planet Mercury the students are supposed to learn about
\begin{itemize}
\item the importance of differential equations in theoretical physics;
\item the numerical implementation of Newtonian dynamics;
\item systematic tests and optimization of computer codes;
\item effective tools to estimate the result a priori as an important cross check;
\item the visualization of numerical results using \vpython{}.
\end{itemize}
The course \todo{what course?} as well as this paper is structured as follows: after an introduction to Newtonian dynamics and the concept of differential equations, their
discretization is discussed based on Newton's law of gravitation and possible extensions thereof.
Afterwards the visualization of the resulting trajectories using \vpython{} is introduced and applied to the problem at hand. In addition,
 tools are developed to extract the relevant quantity from the result of the simulation.
Finally, the principle of dimensional analysis is presented as a tool not only to estimate some effect but also to cross check whether the results of the simulation are sensible.

We are convinced that in order to excite students for numerical simulations it is compulsory to demonstrate their power on an example that catches their interest.
This purpose is served perfectly by the case chosen here, since Einsteins equations of General Relativity are fascinating to a very broad public.
Their detailed study needs a deep knowledge of Differential Geometry. For the study outlined here, however, very little math is necessary, such that high school students from about 10$^{\rm th}$ grade up should benefit from it.
This was already demonstrated in the "Sch\"ulerakademie Teilchenphysik", where this course was tested successfully on two groups consisting in total of 24 German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade in 2015 and 2017\todo{do foreigners know what that means? what are typical ages of students? did grade 13 even exist in 2015?}.

@@

A physical system is said to be understood if the assumed force acting on it leads to the observed trajectories.
In other words, we need to show that we can calculate the location in space of the object of interest at any point in time, once the initial conditions are fixed properly.
If we can neglect the finite size of this object (and in particular its orientation in space), the location is parametrized by a single three vector $\vec r(t)$.
Additionally, we need to be able to calculate the objects velocity $\vec v(t)$ and acceleration $a(t)$ in order to describe and control its dynamics; this will become clear below. The velocity describes a change of location over time which, for some infinitesimally small $\Delta t$, can be written as
\begin{equation}\label{eq:x_update}
\vec r(t+\Delta t) = \vec r(t) + \vec v(t) \Delta t + \ldots \ . \label{eq:vdef}
\end{equation}
Similarly, the acceleration describes a change in velocity, i.e.
\begin{eqnarray}\label{eq:v_update}
\vec v(t+\Delta t) = \vec v(t) + \vec a(t) \Delta t + \ldots\ . \label{eq:adef}
\end{eqnarray}
The dots in these expressions indicate additional terms that may be expressed in terms of higher powers in $\Delta t$. While they are needed in general, for sufficiently small $\Delta t$ they can be safely neglected.
Thus we may define the time derivative via
\begin{equation}
\vec v(t) = \lim_{\Delta t\to 0} \frac{\Delta \vec r(t)}{\Delta t} =: \frac{d\vec r(t)}{d t}  = \dot{\vec  r}(t) \ ,
\end{equation}
where $\Delta \vec r(t)=\vec r(t+\Delta t)-\vec r(t)$ and we introduced a common short hand notation for time derivatives in the last expression.
Analogously we get
\begin{eqnarray}
\vec a(t) &=& \lim_{\Delta t\to 0} \frac{\Delta \vec v(t)}{\Delta t} =: \frac{d\vec v(t)}{dt}  = \dot{\vec v}(t) \ , \\
 &=& \frac{d^2\vec r(t)}{dt^2}  = \ddot{\vec r}(t) \ ,
\end{eqnarray}
where we introduced the second derivative in the second line.

@@

The observed perihelion motion of Mercury is nowadays determined as \todo{reference?}
\[(574.10\pm 0.65)'' \ \mbox{per 100 earth years}\]
where the symbol $''$ denotes "arc seconds": 1$''={(1/3600)}^o$. The bulk of this number
can be understood by the presence of other planets within the Newtonian theory, since
their gravitational force also acts on Mercury. However, a
residual motion of
\begin{equation}
\delta \Theta_M = (43)'' \ \mbox{in 100 earth years} \label{delT}
\end{equation}
remained unexplained, until Einstein quantified the predictions of General Relativity to this particular observable~\cite{Einstein}.

To allow for a movement of the perihelion we need to modify the force that led to Eq.~(\ref{eq:newton}).
This modification should be such that it depends on $r$ and it vanishes for $r\to \infty$.
Based on the discussion presented above we may multiply the right hand side of~(\ref{eq:newton}) with a factor
$(1+\alpha(r_S/r))$, where $\alpha$ denotes some dimensionless parameter.
This seems like a natural way to parametrize the additional potential because it adds the smallest possible deviation from a $1/r$ potential expressed relative to the characteristic length $r_S$.
However, besides $r_S$, which characterizes the gravitational potential of the Sun, there is also a characteristic parameter for the dynamics of Mercury:
\begin{equation}
r_a^2 := \frac{\vec L\,^2}{m^2c^2}= \frac{{(\vec r\times \dot{\vec r} \, )}^2}{c^2} \ , \label{a2def}
\end{equation}
where $\vec L$ denotes the angular momentum, which is a constant of motion for central potentials.
It is easily verified that $r_a^2$ carries dimensions of length squared.
We therefore have to add one more term to the potential; in analogy to the one discussed above, we use $\beta (r_a^2/r^2)$.
Combining everything, we use the following ansatz for the modified equation of motion:
\begin{equation}
\ddot{\vec r} = - \frac{c^2}{2}\frac{r_S}{r^2}\left(1+\alpha\frac{r_S}{r}+\beta\frac{r_a^2}{r^2}\right) \, \frac{\vec{r}}{r} \ .
\label{eq:newton_art}
\end{equation}
For the system at hand, $r_a^2$ may be estimated from the parameters of Mercury at its perihelion: The corresponding velocity is
 $\dot{\vec r}(t=0)=\abs{\vec v_M(0)} = 59$~km/s (here we already indicate that we will start the simulation at the perihelion) and the closest distance between Sun and Mercury is
$ r_{MS}=\abs{\vec r_{MS}(0)} = 46 \cdot 10^6$km~\cite{MercuryFactSheet}. Note that at the perihelion $\vec r$ and $\dot{\vec r}$ are perpendicular to each other.
We thus find that $(r_a^2/ r_{MS}^2) = 4\cdot10^{-8}$. Furthermore $(r_S/r_{MS})=6\cdot 10^{-8}$.
Therefore (for a more detailed discussion about the underlying logic we refer to Sec.~\ref{sec:analysis}) one may expect from both correction terms an effect of similar size. Furthermore, any term of higher order in either $a$ or $r_S$
should be suppressed by additional seven orders of magnitude and thus can be neglected safely.
The parameters $\alpha$ and $\beta$ can be extracted from a fit to data --- this will be done below.
However, since we have only a single number to study, the perihelion shift (see Eq.~(\ref{delT})),
we can only fix either $\alpha$ or $\beta$ (or a linear combination thereof).
The parameters can also be calculated from the underlying theory one uses. The actual values depend on the specific theory;
General relativity gives~\cite{Einstein}
\begin{equation}
	\alpha = 0 \, , \quad \beta = 3 \, . \label{eq:art-prediction}
\end{equation}
Thus we may also use the simulation explained below to calculate the perihelion motion of Mercury from the input
values given in Eq.~(\ref{eq:art-prediction}).

@@

In this section we present the numerical implementation as well as the visualisation of planetary trajectories and in
particular the perihelion motion of Mercury.
We choose \python{}~\cite{Python} as programming language because \python{} is easy to learn, intuitive to understand and an open source language.
No prior knowledge of \python{} or any other programming language is required, as we explain all necessary steps to create the simulation.
In~\ref{appendix:python} we give instructions for setting up \python{} on different operating systems.
We also provide a working example program that can be used as a template \url{this-upload-url}\todo{@arXiv: online link}.

@@

There are several ways to find the position of the perihelion.
The easiest one is to look for the point of minimal distance to the sun --- the definition of the perihelion.
This can be done within the $while$-loop.
To identify a minimal distance to the sun, one has to know the last two positions \code{vec\_rM\_past} and \code{vec\_rM\_before\_last} in addition to the current position.
If the length of the \texttt{last} vector is smaller than that of the \texttt{before\_last} vector and smaller than the current vector, one has passed the perihelion and its position is given by \code{vec\_r\_last}.
It is useful to save the position of the perihelion in a list \code{list\_perih} for a certain number of turns (here e.g.\ \code{max\_turns}=10).
The implementation can be realized as follows\footnote{%
	Note that one stores \code{vec\_r\_last} by making a copy of \code{M.pos}: \code{vec\_r\_last = vector(M.pos)}.
	This is essential because otherwise, if one changes \code{M.pos}, one would automatically change \code{vec\_r\_last} as well.
}
\begin{lstlisting}
# Set up vectors
vec_rM_last = vec_rM0
turns      = 0
max_turns  = 10
list_perih = list()
# Find perihelion for each turn and print it out
while turns < max_turns:
    vec_rM_before_last = vec_rM_last
    # Store position of Mercury in a new vector (since we will change M.pos)
    vec_rM_last = vector(M.pos)
    #<...update Mercury position...>
    # Check if at perihelion
    if (vec_rM_last.mag < M.pos.mag) and (vec_rM_last.mag < vec_rM_before_last.mag):
        list_perih.append(vec_rM_last)
        turns = turns + 1
\end{lstlisting}
The angle between two vectors can be computed using
 \begin{equation}
 	\sphericalangle(\vec{v}_{1},\vec{v}_2) = \mathrm{arccos} \left( \frac{\vec{v}_{1} \cdot \vec{v}_2}{|\vec{v}_{1}|\:|\vec{v}_2|} \right)
	\, .
 \end{equation}
This is readily implemented in \texttt{VPython} via
\begin{lstlisting}
# Import functions to compute angle
from vpython import acos, pi, dot
# Define function for angle extraction
def angle_between(v1, v2):
    return acos( dot(v1, v2) / (v1.mag * v2.mag) ) * 180. / pi
\end{lstlisting}
where the factor $180/\pi$ in the last line converts the unit of the angle from radians into degrees.
To account for the statistical errors (e.g.\ numerical rounding errors) one can average over a few turns.
Depending on the programming proficiency of the students and time constraints, this can be done either by hand or, e.g., by implementing the following code
\begin{lstlisting}
sum_angle=0.
for n in range(1, max_turns):
    # Calculate angle
    sum_angle = sum_angle + angle_between(list_perih[n-1], list_perih[n])
# Display the average
print(sum_angle/(max_turns-1))
\end{lstlisting}
Note that the perihelion motion is computed from the locations stored in \code{list\_perih} and not based on the initial position.
This is important because depending on the initial conditions and numerical uncertainties, the simulation does not necessarily start in the perihelion.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Instructions to install VPython}\label{appendix:python}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In short, to run the above scripts, one needs to install \python{} and \vpython{}.
We recommend the newest \vpython{} version 7 or higher which best runs with the \python{} versions of 3.5 or higher.
It is also possible to use \python{} 2.7 or higher, but the 3.x versions have better \vpython{} support.

For the advanced users, one simply has to update \python{} to these versions and install \vpython{} via the package manager of choice, e.g, \texttt{pip} (see also \cite{VPython}).

Below, we provide detailed descriptions how to install these versions of \python{} and \vpython{} for different operating systems.
Independent of the operating systems, one has to execute the following steps:
\begin{enumerate}
	\item install and update \python{} (at best 3.x, x$\geq5$);
	\item install and update a \python{} package manager;
	\item install \vpython{} and dependencies.
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Linux based systems (including macOS)}\label{appendix:python-max}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
All Linux based system come with an installation of \python{}.
Thus, if not already at the newest version, one just has to update existing installations.
This is best done by using a distribution manager.
Native to Linux based systems is for example \texttt{dpkg} and \texttt{apt-get}, or \texttt{homebrew} and \texttt{MacPorts} for macOS.
While \texttt{dpkg} and \texttt{apt-get} usually come with the linux installations, macOS user must install such distribution managers at first.
For this we refer to \cite{MacPorts} or \cite{Homebrew}.

As the start one should open a terminal/console (bash).
Depending on the installation, one usually needs superuser rights in order to update \python{}.
Therefore one has to prepend \code{sudo} to the following commands and confirm them by entering the superuser password.
The installations using a distribution manager are save and tested by the vast community.
In the following, the successive order of commands to update \python{} is presented for \texttt{apt-get} (Linux) and \texttt{MacPorts} (macOS), but are most similar for other distribution managers.
\begin{table}[h!]
\centering
\footnotesize
\begin{tabular}{l || l | l}
	& \texttt{apt-get} & \texttt{MacPorts} \\\hline
	1. Update the distribution manager & \code{sudo apt-get update} & \code{sudo port selfupdate} \\
	2. Install \& update \python{} & \code{sudo apt-get install python3} & \code{sudo port install python36} \\
	3. Installing \python{} package manager & \code{sudo apt-get install python3-pip} & \code{sudo port install py36-pip}
\end{tabular}
\normalsize
\end{table}

Depending on the previously installed packages, this command might install further dependencies automatically (after confirmation).
One should test if the installation was successful by typing \code{python3 --version} or \code{python3.6 --version} (depending on the installation name) which prints out the version of the \python{} installation (and should be larger than 3.5).
Next one must update the package manager to get the newest versions of \python{} packages.
This is done by running
\begin{table}[h!]
\centering
\footnotesize
\begin{tabular}{l || l | l}
	& \texttt{apt-get} & \texttt{MacPorts} \\\hline
	4. Update \code{pip} & \code{pip3 install --upgrade --user pip wheel} & \code{pip-3.6 install --upgrade --user pip wheel} \\
\end{tabular}
\normalsize
\end{table}\\
Note that this just upgrades the packages for the current user.
If one wants the systemwide installation, one again has to prepend \code{sudo} and can ignore the \code{--user} flag (\code{sudo pip3 install --upgrade pip wheel}).
Furthermore, depending on the installed version of \code{pip}, it could have a different name.
This is best figured out by typing \code{pip} and pressing \texttt{tab}.

Finally, one can install \vpython{} by running
\begin{table}[h!]
\centering
\footnotesize
\begin{tabular}{l || l | l}
	& \texttt{apt-get} & \texttt{MacPorts} \\\hline
	5. Install \vpython{} & \code{pip3 install --user vpython} & \code{pip-3.6 install --user vpython} \\
\end{tabular}
\normalsize
\end{table}

At this point, everything is installed and one can start the simulation.
To test the \vpython{} installation one should run \code{python3} or \code{python3.6}, which starts a \python{} console.
Next, one shall run \code{from vpython import *}, which loads the \vpython{} package and opens the default browser.
Running \code{box()} should generate a white box in the browser window, which can be rotated using the mouse.
The program is quit by running \code{quit()} or pressing \texttt{Ctrl-c}.

In general, there are multiple ways to run \python{} scripts.
For the sake of this paper, we suggest two scenarios.
\begin{enumerate}
\item The first scenario to run \python{} scripts is that one writes and saves a \python{} file, e.g., \texttt{test.py}, which can be edited using any text editor of choice.
	To run the \python{} script one must navigate the terminal to the containing folder using the \code{cd} (and \code{ls}) command.
	While \code{ls} prints out the folders and files in the current directory, \code{cd} changes the directory.
	The command \code{cd} either takes relative paths (directories which are in the current directory), e.g., \code{cd folder}, or absolute paths beginning with a \code{/}, e.g., \code{cd /Users/name/Desktop/}.
	Once one has navigated to folder which contains the \python{} file --- e.g., the file name is listed when typing \code{ls} --- than one can type \code{python3 test.py} or \code{python3.6 test.py} to execute the script.
	Again, this script can be ended (if it does not end automatically) by pressing \texttt{Ctrl-c}.
\item The second scenario to run \python{} scripts is given by the interactive \texttt{ipython} or \texttt{jupyter} notebooks.
	This feature is automatically installed when installing the newest version of \vpython{}.
	To start an interactive session, one has to run \code{jupyter notebook} (or equivalent, depending on the installation name. Again, this can be figured out by typing \code{jupyter} and pressing \texttt{tab}).
	This will print a link (containing the name \texttt{localhost}) and open the browser at the link-address.
	Within this browser window, one can freely navigate to the directory of choice and create interactive \python{} and \vpython{} sessions by clicking on existing files or creating new ones (upper right button \texttt{new}$\rightarrow$\vpython{}).
\end{enumerate}
We provide example files for both scenarios.\todo{@arXiv: online link}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Windows}\label{appendix:python-windows}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

As the first step, one must install \python{}, again there are several possibilities but we suggest \texttt{Anaconda} for an intuitive installation.
Following the link \cite{conda}, one has two possible (free) options, \texttt{Anaconda} and \texttt{Miniconda}.
Both are equally sufficient.
One can install further updates after the initial installation.
In the case of \texttt{Miniconda} \cite{Miniconda} one shall download the newest \python{} 3.x (x $\geq5$) installer for windows and the right system type (32-bit or 64-bit).
The system type can be found by pressing \texttt{Windows+I}, navigating to \texttt{System}$\rightarrow$\texttt{About} and looking at the \texttt{System type} entry.

Installing one of the two executables creates a new \texttt{Windows} application called \texttt{Anaconda Prompt}.
Starting this application opens a terminal which will be used to install the required \python{} packages.
At first, one should test if the \texttt{Anaconda} installation was successful by running the command \code{conda list}.
This command should print a list of already installed packages.
Also one should run \code{conda update conda} to make sure one has the most recent installation.

To install \vpython{} one simply has to run \code{conda install -c vpython vpython}.
The additional flag \code{-c vpython} tells \texttt{Anaconda} the "channel" where to look for the \vpython{} package.
Running this command prints out a list of dependencies which will be automatically installed in order for \vpython{} to work --- this must be confirmed.

In principle this is already sufficient to run \vpython{} scripts.
One could furthermore install \texttt{Anaconda}s native navigator \texttt{anaconda-navigator} by running \code{conda install anaconda-navigator}.
This will install another application, \texttt{Anaconda Navigator}, which can be run from the Desktop and makes it easier to program in \python{}.

Again there are several possibilities to run \python{} files on \texttt{Windows}.
\begin{itemize}
	\item One could create a file \texttt{test.py}, edit it with the editor of choice and specify that this file should be opened with \texttt{Python3}.
		To do so, \texttt{right click} on  the file, select \texttt{open with} and locate the \texttt{Anaconda Python3} installation.
		Programs can be closed by closing the running window.
	\item One can run \code{python3} in the \texttt{Anaconda Prompt} application to start an interactive session.
		Programs can be closed by closing the running window.
	\item Start the \texttt{Anaconda Navigator} application, open an interactive \texttt{qtconsole} and program within the newly opened terminal.
		Programs can be closed by closing the running window.
	\item Start the \texttt{Anaconda Navigator} application and click on the \texttt{jupyter notebook} button.
		This opens an interactive \python{} session in the default web-browser.
		Here, one can open existing notebooks or create new files (upper right button \texttt{new}$\rightarrow$\vpython{}).
\end{itemize}

To start and test the \vpython{} installation one should pick any out of these options and execute the code
\begin{lstlisting}
from vpython import *
box()
\end{lstlisting}
Running this code should open the default browser and render a white box on black background.
Using the mouse (right click and drag), one should be able to rotate the box.
It was found that the \texttt{Internet Explorer} is not able to properly render the animation.
Here we suggest to use the \texttt{Firefox} web-browser as the default choice.

Once this test case works, one can start to test the Mercury orbit simulation.
We provide example files for both \texttt{juypter notebooks} and regular \python{} files.

\todo{@arXiv: online link}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage
\section{The complete code}\label{sec:code}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Below we provide the complete code as developed above.
\lstinputlisting[language=Python]{mercury_orbit_base.py}
Further examples and template files can be found online\cite{}.\todo[inline]{@arXiv: online link}

>>>>

Numerical simulations play a key role in modern physics for they allow one to tackle theoretical problems not accessible otherwise. This might be the case because there are too many particles participating in the system (as in simulations for weather predictions) or the interactions are too complicated to allow for a systematic, perturbative approach (as in theoretical descriptions of nuclear particles at the fundamental level).
This paper introduces a project that allows one to demonstrate the power of numerical simulations.
On the example of the perihelion motion of the planet Mercury the students are supposed to learn about
\begin{itemize}
\item the importance of differential equations in theoretical physics;
\item the numerical implementation of Newtonian dynamics;
\item systematic tests and optimization of computer codes;
\item effective tools to estimate the result a priori as an important cross check;
\item the visualization of numerical results using \vpython{}.
\end{itemize}
We are convinced that in order to excite students for numerical simulations it is compulsory to demonstrate their power on an example that catches their interest.
This purpose is served perfectly by the case chosen here, since Einsteins equations of General Relativity are fascinating to a very broad public.
Their detailed study needs a deep knowledge of Differential Geometry. For the study outlined here, however, very little math is necessary, such that high school students familiar to derivatives benefit from it.
This was already demonstrated in the "Sch\"ulerakademie Teilchenphysik", where this course was tested successfully on two groups consisting in total of 24 German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade in 2015 and 2017.

The course as well as this paper is structured as follows: after an introduction to Newtonian dynamics and the concept of differential equations, their
discretization is discussed based on Newton's law of gravitation and possible extensions thereof.
Afterwards the visualization of the resulting trajectories using \vpython{} is introduced and applied to the problem at hand. In addition,
 tools are developed to extract the relevant quantity from the result of the simulation.
Finally, the principle of dimensional analysis is presented as a tool not only to estimate some effect but also to cross check whether the results of the simulation are sensible.

@@

A physical system is said to be understood if the assumed forces acting on it lead to the observed trajectories.
In other words, we need to show that we can calculate the location in space of the object of interest at any point in time, once the initial conditions are fixed properly.
If we can neglect the finite size of this object (and in particular its orientation in space), the location is parametrized by a single three vector $\vec r(t)$.
Additionally, we need to be able to calculate the objects velocity $\vec v(t)$ and acceleration $a(t)$ in order to describe and control its dynamics; this will become clear below. The velocity describes a change of location over time which, for some infinitesimally small $\Delta t$, can be written as
\begin{equation}\label{eq:x_update}
\vec r(t+\Delta t) = \vec r(t) + \vec v(t) \Delta t + \ldots \ . \label{eq:vdef}
\end{equation}
Similarly, the acceleration describes a change in velocity, i.e.
\begin{eqnarray}\label{eq:v_update}
\vec v(t+\Delta t) = \vec v(t) + \vec a(t) \Delta t + \ldots\ . \label{eq:adef}
\end{eqnarray}
The dots in these expressions indicate additional terms that may be expressed in terms of higher powers in $\Delta t$. While they are needed in general, for sufficiently small $\Delta t$ they can be safely neglected.
Thus we may define the time derivative via
\begin{equation}
\vec v(t) = \lim_{\Delta t\to 0} \frac{\Delta \vec r(t)}{\Delta t} =: \frac{d\vec r(t)}{d t}  = \dot{\vec  r}(t) \ ,
\end{equation}
where $\Delta \vec r(t)=\vec r(t+\Delta t)-\vec r(t)$ and we introduced a common short hand notation for time derivatives in the last expression.
Analogously we get
\begin{eqnarray}
\vec a(t) &=& \lim_{\Delta t\to 0} \frac{\Delta \vec v(t)}{\Delta t} =: \frac{d\vec v(t)}{dt}  = \dot{\vec v}(t) \ , \\
 &=& \frac{d^2\vec r(t)}{dt^2}  = \ddot{\vec r}(t) \ ,
\end{eqnarray}
where we introduced the second derivative in the second line.

@@

The observed perihelion motion of Mercury is nowadays determined as
\[(574.10\pm 0.65)'' \ \mbox{per 100 earth years \cite{RevModPhys.19.361}}\]
where the symbol $''$ denotes "arc seconds": 1$''={(1/3600)}^o$. The bulk of this number
can be understood by the presence of other planets within the Newtonian theory, since
their gravitational force also acts on Mercury. However, a
residual motion of
\begin{equation}
\delta \Theta_M = (43)'' \ \mbox{in 100 earth years} \label{delT}
\end{equation}
remained unexplained, until Einstein quantified the predictions of General Relativity to this particular observable~\cite{Einstein}.

To allow for a movement of the perihelion we need to modify the force that led to Eq.~(\ref{eq:newton}).
This modification should be such that it depends on $r$ and it vanishes for $r\to \infty$.
Based on the discussion presented above we may multiply the right hand side of~(\ref{eq:newton}) with a factor
$(1+\alpha(r_S/r))$, where $\alpha$ denotes some dimensionless parameter.
This seems like a natural way to parametrize the additional potential because it adds the smallest possible deviation from an $1/r$ potential expressed relative to the characteristic length $r_S$.
However, besides $r_S$, which characterizes the gravitational potential of the Sun, there is also a characteristic parameter for the dynamics of Mercury:
\begin{equation}
r_a^2 := \frac{\vec L\,^2}{m^2c^2}= \frac{{(\vec r\times \dot{\vec r} \, )}^2}{c^2} \ , \label{a2def}
\end{equation}
where $\vec L$ denotes the angular momentum, which is a constant of motion for central potentials.
It is easily verified that $r_a^2$ carries dimensions of length squared.
We therefore have to add one more term to the potential; in analogy to the one discussed above, we use $\beta (r_a^2/r^2)$.
Combining everything, we use the following ansatz for the modified equation of motion:
\begin{equation}
\ddot{\vec r} = - \frac{c^2}{2}\frac{r_S}{r^2}\left(1+\alpha\frac{r_S}{r}+\beta\frac{r_a^2}{r^2}\right) \, \frac{\vec{r}}{r} \ .
\label{eq:newton_art}
\end{equation}
For the system at hand, $r_a^2$ may be estimated from the parameters of Mercury at its perihelion: The corresponding velocity is
 $\dot{\vec r}(t=0)=\abs{\vec v_M(0)} = 59$~km/s (here we already indicate that we will start the simulation at the perihelion) and the closest distance between Sun and Mercury is
$ r_{MS}=\abs{\vec r_{MS}(0)} = 46 \cdot 10^6$km~\cite{MercuryFactSheet}. Note that at the perihelion $\vec r$ and $\dot{\vec r}$ are perpendicular to each other.
We thus find that $(r_a^2/ r_{MS}^2) = 4\cdot10^{-8}$. Furthermore $(r_S/r_{MS})=6\cdot 10^{-8}$.
Therefore (for a more detailed discussion about the underlying logic we refer to Sec.~\ref{sec:analysis}) one may expect from both correction terms an effect of similar size. Furthermore, any term of higher order in either $a$ or $r_S$
should be suppressed by additional seven orders of magnitude and thus can be neglected safely.
The parameters $\alpha$ and $\beta$ can be extracted from a fit to data --- this will be done below.
However, since we have only a single number to study, the perihelion shift (see Eq.~(\ref{delT})),
we can only fix either $\alpha$ or $\beta$ (or a linear combination thereof).
The parameters can also be calculated from the underlying theory one uses. The actual values depend on the specific theory;
General relativity gives~\cite{Einstein}
\begin{equation}
	\alpha = 0 \, , \quad \beta = 3 \, . \label{eq:art-prediction}
\end{equation}
Thus we may also use the simulation explained below to calculate the perihelion motion of Mercury from the input
values given in Eq.~(\ref{eq:art-prediction}).

@@

In this section we present the numerical implementation as well as the visualization of planetary trajectories and in
particular the perihelion motion of Mercury.
We choose \python{}~\cite{Python} as programming language because \python{} is easy to learn, intuitive to understand and an open source language.
No prior knowledge of \python{} or any other programming language is required, as we explain all necessary steps to create the simulation.
We provide online instructions for setting up \python{} and \vpython{} on different operating systems \cite{scripts}.
In \ref{sec:code}, we also provide a working example as well as possible extensions and template files online.

@@

There are several ways to find the position of the perihelion.
The easiest one is to look for the point of minimal distance to the sun --- the definition of the perihelion.
This can be done within the $while$-loop.
To identify a minimal distance to the sun, one has to know the last two positions \code{vec\_rM\_past} and \code{vec\_rM\_before\_last} in addition to the current position.
If the length of the \texttt{last} vector is smaller than that of the \texttt{before\_last} vector and smaller than the current vector, one has passed the perihelion and its position is given by \code{vec\_r\_last}.
It is useful to save the position of the perihelion in a list \code{list\_perih} for a certain number of turns (here e.g.\ \code{max\_turns = 10}).
The implementation can be realized as follows\footnote{%
	Note that one stores \code{vec\_r\_last} by making a copy of \code{M.pos}: \code{vec\_r\_last = vector(M.pos)}.
	This is essential because otherwise, if one changes \code{M.pos}, one would automatically change \code{vec\_r\_last} as well.
}
\begin{lstlisting}
# Set up vectors
vec_rM_last = vec_rM0
turns      = 0
max_turns  = 10
list_perih = list()
# Find perihelion for each turn and print it out
while turns < max_turns:
    vec_rM_before_last = vec_rM_last
    # Store position of Mercury in a new vector (since we will change M.pos)
    vec_rM_last = vector(M.pos)
    #<...update Mercury position...>
    # Check if at perihelion
    if (vec_rM_last.mag < M.pos.mag) and (vec_rM_last.mag < vec_rM_before_last.mag):
        list_perih.append(vec_rM_last)
        turns = turns + 1
\end{lstlisting}
The angle between two vectors can be computed using
 \begin{equation}
 	\sphericalangle(\vec{v}_{1},\vec{v}_2) = \mathrm{arccos} \left( \frac{\vec{v}_{1} \cdot \vec{v}_2}{|\vec{v}_{1}|\:|\vec{v}_2|} \right)
	\, .
 \end{equation}
This is readily implemented in \texttt{VPython} via
\begin{lstlisting}
# Import functions to compute angle
from vpython import acos, pi, dot
# Define function for angle extraction
def angle_between(v1, v2):
    return acos( dot(v1, v2) / (v1.mag * v2.mag) ) * 180. / pi
\end{lstlisting}
where the factor $180/\pi$ in the last line converts the unit of the angle from radians into degrees.
To account for the statistical errors (e.g.\ numerical rounding errors) one can average over a few turns.
Depending on the programming proficiency of the students and time constraints, this can be done either by hand or, e.g., by implementing the following code
\begin{lstlisting}
sum_angle=0.
for n in range(1, max_turns):
    # Calculate angle
    sum_angle = sum_angle + angle_between(list_perih[n-1], list_perih[n])
# Display the average
print(sum_angle/(max_turns-1))
\end{lstlisting}
Note that the perihelion motion is computed from the locations stored in \code{list\_perih} and not based on the initial position.
This is important because depending on the initial conditions and numerical uncertainties, the simulation does not necessarily start in the perihelion.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage
\section{The complete code}\label{sec:code}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Below we provide the complete code as developed above.
\lstinputlisting[language=Python,basicstyle=\scriptsize]{mercury_orbit_base.py}
Further examples and template files can be found online \cite{scripts}.
